A separation-of-variable dynamic stiffness matrix method for the free vibration of orthotropic rectangular thin plates

### 2025-07-25 Paper7.tex
Editor

\documentclass[preprint,12pt]{elsarticle}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{array}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{subcaption}
\usepackage{mwe}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage[noabbrev,nameinlink]{cleveref}
\usepackage[title]{appendix}
\usepackage[utf8]{inputenc}
\usepackage{stanli}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{xspace}
\usepackage{lmodern}
\usepackage{nomencl}
\makenomenclature
\usepackage{longtable}
\usepackage{lineno}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{shapes,positioning}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\newcommand{\kph}{~\unit{\kilo\metre\per\hour}\xspace}
\newcommand{\um}{~\unit{\metre}\xspace}
\newcommand{\ukg}{~\unit{\kilogram}\xspace}
\newcommand{\urho}{~\unit{\kilogram\per\meter}\xspace}
\newcommand{\uI}{~\unit{\kilogram\square\metre}\xspace}
\newcommand{\urd}{~\unit{\newton\square\metre}\xspace}
\newcommand{\ustif}{~\unit{\newton\per\metre}\xspace}
\newcommand{\udamp}{~\unit{\newton\second\per\metre}\xspace}
\newcommand{\DAF}{\textnormal{DAF}}
%\newcommand{\ii}{\textnormal{i}}
\newcommand{\id}{\mathrm{d}\mkern1mu}
\newcommand{\ii}{\mathrm{i}\mkern1mu}
\newcommand{\nukg}{~\unit{\kilogram}\xspace}
\newcommand{\nustif}{~\unit{\newton\per\metre}\xspace}
\newcommand{\nudamp}{~\unit{\newton\second\per\metre}\xspace}
\newcommand{\nugpa}{~\si{\giga\pascal}\xspace}
\begin{document}

\begin{frontmatter}
\title{Dynamic stiffness matrix based on the extended separation-of-variables solution for the free vibration of orthotropic rectangular thin plates}
\author[1]{Shiyi Mei}\ead{shiyi.mei1@monash.edu}
\author[1]{Colin Caprani\corref{cor1}}\ead{colin.caprani@monash.edu}
\author[2]{Daniel Cantero}\ead{daniel.cantero@ntnu.no}
\cortext[cor1]{Corresponding author}
\affiliation[1]{organization={Department of Civil Engineering, Monash University},city={Melbourne},state={Victoria},country={Australia}}
\affiliation[2]{organization={Department of Structural Engineering, Norwegian University of Science \& Technology NTNU},city={Trondheim}, country={Norway}}
\begin{abstract}
The dynamic stiffness matrix (DSM) based on the extended separation-of-variables (SOV) solution is developed for the free vibration analysis of an orthotropic rectangular thin plate with general homogeneous boundary conditions.  
The method combines the advantages of the DSM method and the SOV method.
The SOV solution satisfies the governing differential equation derived from Rayleigh’s principle and is used to formulate dynamic stiffness matrices.
Owing to the characteristics of the SOV solution, the fully clamped boundary condition problem associated with the Wittrick–Williams algorithm is resolved. 
The enhanced algorithm is further proposed to solve dynamic stiffness matrices, rather than solving eigenvalue equations.
A numerical technique for mode shape computation is also introduced.
The accuracy of the proposed method is validated through numerical experiments.

\end{abstract}
\end{frontmatter}
\linenumbers
\section{Introduction}
Rectangular plates play an important role in various engineering fields, including civil, mechanical, and aerospace engineering \citep{biancolini2005approximate}. 
The free vibration of plates has been a fundamental research problem for over two centuries. 
The earliest exact solutions for this problem are the Navier \citep{navier1823extrait} and Levy \citep{levy1899equilibre} solutions, which require at least one pair of opposite edges to be simply supported or guided.
To solve problems with other boundary conditions, approximate solutions such as the Rayleigh–Ritz method \cite{leissa1973free} and the Galerkin method \cite{laura1967study} have been widely applied. 
For these approximation methods, beam functions, polynomials, trigonometric functions, and their combinations \cite{li2004vibration} are commonly used as the assumed approximate functions.
The accuracy of these solutions depends on how well the assumed approximate functions represent the displacement of the plate.

Besides the approximation methods, several analytical methods have been developed over past decades, including the Kantorovich-Krylov method \cite{kantorovich1958approximate,kerr1968extension}, the symplectic eigenfunction expansion method \cite{zhong1995new,xing2009new2}, the separation-of-variable (SOV) method \cite{xing2009new}, the dynamic stiffness matrix (DSM) method \cite{banerjee1997dynamic}, and series expansion-based methods \cite{xing2022review}.
The series expansion-based methods include the superposition method \cite{timoshenko1940theory,gorman2005free}, Fourier series method \cite{khov2009accurate,li2009exact}, the finite integral transform method \cite{li2009finite,zhong2013free}, and other series methods. 
These methods represent the plate displacement in terms of an infinite series and mostly are capable of handling any general boundary conditions. 
However, sufficient truncation of the series is required to ensure the accuracy and convergence of the results, and the eigenvalue equation is generally difficult to express explicitly.
Therefore, solving the corresponding eigenvalue problem can be computationally expensive.

Despite being a powerful method for the dynamic analysis of plate assemblies, the finite element method (FEM) requires a sufficient number of elements and is computationally expensive to accurately capture higher-order modes.
Thus, the DSM method was developed as an accurate and efficient analytical approach to alternatively solve complex plate structures \cite{boscolo2011dynamic,fazzolari2013exact}. 
The DSM can be considered as an analytical FEM since the mode functions of the plate are expressed by analytical solutions, where Levy-type solution \cite{ghorbel2015dynamic} or components of infinite Fourier series \cite{banerjee2015dynamic,liu2016free} are applied.
To avoid solving the cumbersome transcendental frequency equation directly, the Wittrick-Williams (W-W) algorithm \cite{wittrick1971general} is applied to the eigenvalue problem.
The W-W algorithm determines the lower and upper bounds of natural frequencies to arbitrary precision rather than solving the frequency equation directly. 
Thus, the DSM has the potential to be effectively and systematically solved using the W-W algorithm.
However, a critical part in applying the W-W algorithm is to determine all natural frequencies of the fully clamped structure within the interested frequency range, \textit{a priori}.
Strategies such as using a sufficiently fine mesh or including a sufficient number of terms in series expansions \cite{banerjee2015dynamic} can ensure that all fully clamped frequencies are accounted for, thereby maintaining the accuracy of the algorithm. 
However, these approaches are computationally expensive and complex, posing a significant obstacle to the wider adoption and application of the DSM method based on the W-W algorithm \cite{han2018improved}.
To resolve the fully clamped plate problem, \citet{liu2015exact} suggested that the frequencies can be indirectly obtained from the simply supported plate problem, where the Navier solution serves as the analytical solution. 
This provides a significant enhancement to the W-W algorithm, increasing the efficiency of applying DSM methods. 
However, the solutions are not explicit and closed-form, but are expressed in an infinite series form, where a sufficient number of truncation terms is required to ensure accuracy.

Inspired by the Navier and Levy solutions, \citet{xing2009new} proposed the SOV method, which provides concise and explicit eigensolutions.
The mode shape function has a separable form, $\phi(x) \psi(y)$, requiring only one $\phi(x)$ and one $\psi(y)$ for each mode order, allowing each eigenvalue equation to be explicitly expressed.
However, this SOV method is not suitable to deal with plates with free boundary conditions.
Therefore, an extended SOV method \cite{xing2018overall, xing2020extended} based on the Rayleigh quotient was proposed to accommodate plates with all four classical boundary conditions, i.e., simply supported, clamped, guided, and free.
Based on the Rayleigh quotient model, alternative iterative and improved SOV methods have been subsequently proposed \cite{xing2020improved}.
Although SOV methods provide concise closed-form analytical solutions, they require solving a specific set of highly nonlinear eigenvalue equations for each type of boundary condition.
However, even when considering only the four classic homogeneous cases, it becomes evident that 55 different boundary condition combinations exist for a rectangular plate, making the process tedious.

In this study, the SOV method is further extended to analyze the vibrations of plates with elastically restrained edges.
This extended SOV solution is then employed to construct dynamic stiffness matrices, which accommodate all general homogeneous boundary conditions.
By taking advantage of both the SOV and DSM methods, an enhanced W-W algorithm is developed to solve the eigenvalue problem without directly solving the eigenvalue equations.
This enhanced approach resolves the challenge of determining fully clamped frequencies, a well-known limitation in the application of the  W-W algorithm.
In addition, a novel numerical technique is proposed to compute the mode shape coefficients.

\FloatBarrier
\section{Mathematical model}\label{sec:Mathematical model}
\begin{figure}[!htbp]
	\centering
	\resizebox{0.8\textwidth}{!}
	{
		\input{figs/general_plate_tikz.tex}
	}
	\caption{\small The orthotropic rectangular plate with all edges elastically restrained.} 
	\label{fig:platemode}
\end{figure}
Consider a thin orthotropic rectangular plate of length $2a$ and width $2b$, with all four edges restrained by vertical translational springs $k^v$ and rotational springs $k^r$, as shown in \Cref{fig:platemode}. 
The coordinate origin is located at the center of the plate.

@@

Following the same steps as for the x-direction, the magnitude of potential energy of the plate in the $y$-direction can be obtained as:
%
\begin{equation}\label{eq:poten_energy_toly2}
	\begin{split}
		U
		&= \frac{ab}{2} \int_{-1}^{1} \left[ \frac{D_{11}}{a^4} J_4 \psi^2 
		+ \frac{2D_{12}}{a^2b^2} J_2 \frac{\id^2\psi }{\id \eta^2} \psi 
		+ \frac{D_{22}}{b^4} J_1 \left(\frac{\id^2\psi }{\id \eta^2} \right)^2 \right. \\
		&\qquad \left. + \frac{4D_{66}}{a^2b^2} J_3 \left(\frac{\id \psi }{\id \eta} \right)^2 \right] \, \id\eta  + abJ_1 \left[ \frac{k^r_{\eta=1}}{b^3} \left( \frac{\id \psi}{\id \eta} \right)^2
		+ \frac{k^v_{\eta=1}}{b}  \left(\psi\right)^2 \right]_{\eta=1} \\
		&\qquad + abJ_1 \left[ \frac{k^r_{\eta=-1}}{b^3} \left( \frac{\id \psi}{\id \eta} \right)^2
		+ \frac{k^v_{\eta=-1}}{b} \left(\psi\right)^2 \right]_{\eta=-1},
	\end{split}
\end{equation}
%
where the integral parameters $J_1$, $J_2$, $J_3$, and $J_4$ are defined and expressed in \Cref{sec:appendA}.

The coefficient $T_0$ of the kinetic energy from \Cref{eq:kine_energy_mag} for the plate in the $y$-direction can be expressed as:

@@

\begin{equation}\label{eq:kine_energyy}

@@

		T_0 &= \frac{ab}{2} \rho h J_1 \int_{-1}^{1} \psi^2 \, \id\eta.

@@

Take the Rayleigh principle in the form:
		\delta U &= \omega_y^2 \, \delta T_0.
By substituting \Cref{eq:poten_energy_toly2,eq:kine_energyy} into \Cref{eq:Rayleighy}, and relieving $ \delta \psi $ and $\delta\frac{\id \psi}{\id \eta}$ in \Cref{eq:Rayleighy} by calculus of variation, yields:
\begin{equation}\label{eq:Rayleighy2}
		0 &= \int_{-1}^{1} \left[ \frac{D_{22}}{b^4} J_1 \frac{d^4 \psi}{\id \eta^4} 
		+ \left( \frac{2D_{12}}{a^2b^2} J_2 - \frac{4D_{66}}{a^2b^2} J_3 \right) \frac{\id^2 \psi}{\id \eta^2} \right. \\
		&\qquad \left. + \left( \frac{D_{11}}{a^4} J_4 - \omega_y^2 \rho h J_1 \right) \psi \right] \delta \psi \, \id\eta \\
		&\qquad + \frac{2k^v_{\eta=-1}}{b} J_1 (\psi \delta \psi)_{\eta=-1} 
		+ \frac{2k^v_{\eta=1}}{b} J_1 (\psi \delta \psi)_{\eta=1} \\
		&\qquad + \left[ \left( \frac{4D_{66}}{a^2b^2} J_3 - \frac{D_{12}}{a^2b^2} J_2 \right) \frac{\id \psi}{\id \eta} 
		- \frac{D_{22}}{b^4} J_1 \frac{\id^3 \psi}{\id \eta^3} \right] \delta \psi \Big|^{\eta=1}_{\eta=-1} \\
		&\qquad + \left( \frac{D_{12}}{a^2b^2} J_2 \psi + \frac{D_{22}}{b^4} J_1 \frac{\id^2 \psi}{\id \eta^2} \right) 
		\delta\frac{\id  \psi}{\id \eta} \Big|^{\eta=1}_{\eta=-1} \\
		&\qquad + \frac{2k^r_{\eta=-1}}{b^3} J_1 \left( \frac{\id \psi}{\id \eta} \delta\frac{\id \psi}{\id \eta} \right)_{\eta=-1} 
		+ \frac{2k^r_{\eta=1}}{b^3} J_1 \left( \frac{\id \psi}{\id \eta} \delta\frac{\id \psi}{\id \eta} \right)_{\eta=1}.
Thus, the governing differential equation in the $y$-direction can be obtained from the integration part in \Cref{eq:Rayleighy2}:

@@

		\alpha_2 &= \frac{1}{\chi} \sqrt{ \sqrt{ \left( \frac{D_{12}J_2}{D_{22}J_1} - 2\frac{D_{66}J_3}{D_{22}J_1} \right)^2 - \frac{D_{11}J_4}{D_{22}J_1} + \frac{a^4 D_{11}}{D_{22}} \Omega_y^4 } + \frac{D_{12}J_2}{D_{22}J_1} - 2 \frac{D_{66}J_3}{D_{22}J_1}}, \\
		\beta_2 &= \frac{1}{\chi} \sqrt{ \sqrt{ \left( \frac{D_{12}J_2}{D_{22}J_1} - 2 \frac{D_{66}J_3}{D_{22}J_1} \right)^2 - \frac{D_{11}J_4}{D_{22}J_1} + \frac{a^4 D_{11}}{D_{22}} \Omega_y^4 } - \frac{D_{12}J_2}{D_{22}J_1} + 2 \frac{D_{66}J_3}{D_{22}J_1}}.

@@

The boundary conditions along the edges in the $y$-direction can be obtained from the remaining $\delta \psi$ and $\delta \frac{\id \psi}{\id \eta}$ parts in \Cref{eq:Rayleighy2}. 
The shear force equilibrium can be obtained from the $\delta \psi$ part:
\begin{equation}\label{eq:inertial_forcey}
		& \left[ \left( \frac{4D_{66}}{a^2b^2} J_3 - \frac{D_{12}}{a^2b^2} J_2 \right) \frac{\id \psi}{\id \eta} 
		- \frac{D_{22}}{b^4} J_1 \frac{\id^3 \psi}{\id \eta^3} \right] \Big|^{\eta=1}_{\eta=-1} \\
		&\qquad + \frac{2k^v_{\eta=-1}}{b} J_1 (\psi )_{\eta=-1} + \frac{2k^v_{\eta=1}}{b} J_1 (\psi )_{\eta=1} = 0,
and from the $\delta \frac{\id \psi}{\id \eta}$ part, the bending moment equilibrium:
\begin{equation}\label{eq:inertial_forcey2}	
 	\left( \frac{D_{12}}{a^2b^2} J_2 \psi + \frac{D_{22}}{b^4} J_1 \frac{\id^2 \psi}{\id \eta^2} \right) 
 	\Big|^{\eta=1}_{\eta=-1} 
 	&+ \frac{2k^r_{\eta=-1}}{b^3} J_1 \left( \frac{\id \psi}{\id \eta} \right)_{\eta=-1} \\
 	&\qquad+ \frac{2k^r_{\eta=1}}{b^3} J_1 \left( \frac{\id \psi}{\id \eta} \right)_{\eta=1}=0.

@@

\section{Frequency and mode shape computation}\label{sec:WWA}
\subsection{Wittrick–Williams algorithm and enhancement}\label{sec:WWA2}
The Wittrick–Williams (W-W) algorithm \cite{wittrick1971general} is an effective method for determining the natural frequencies from the dynamic stiffness matrix with high reliability. 
Instead of directly solving the equations, the algorithm computes the total number $J$ of natural frequencies below a given frequency $\omega^*$, which is represented as:
\begin{equation}\label{eq:WWalgorithm}
	J(\omega^*) = J_0(\omega^*) + s\{\mathbf{K}^{\Delta}(\omega^*)\} = J_0(\omega^*) + J_k(\omega^*),
\end{equation}
where $J_0$ represents the number of natural frequencies of the structure with all ends fully clamped, $\mathbf{K}^{\Delta}$ is the upper triangular matrix obtained from the dynamic stiffness matrix $\mathbf{K}$ after applying Gaussian elimination, and $J_k(\omega^*)$ denotes the number of negative elements in the leading diagonal of $\mathbf{K}^{\Delta}$.

>>>>

\documentclass[preprint,12pt,number]{elsarticle}
\usepackage[percent]{overpic} % in your preamble
\usepackage{enumitem}
\usepackage{setspace} % for line spacing
\biboptions{numbers,sort&compress}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{array}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{subcaption}
\usepackage{mwe}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage[noabbrev,nameinlink]{cleveref}
\usepackage[title]{appendix}
\usepackage[utf8]{inputenc}
\usepackage{stanli}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{xspace}
\usepackage{lmodern}
\usepackage{nomencl}
\makenomenclature
\usepackage{longtable}
\usepackage{lineno}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{shapes,positioning}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\newcommand{\kph}{~\unit{\kilo\metre\per\hour}\xspace}
\newcommand{\um}{~\unit{\metre}\xspace}
\newcommand{\ukg}{~\unit{\kilogram}\xspace}
\newcommand{\urho}{~\unit{\kilogram\per\meter}\xspace}
\newcommand{\uI}{~\unit{\kilogram\square\metre}\xspace}
\newcommand{\urd}{~\unit{\newton\square\metre}\xspace}
\newcommand{\ustif}{~\unit{\newton\per\metre}\xspace}
\newcommand{\udamp}{~\unit{\newton\second\per\metre}\xspace}
\newcommand{\DAF}{\textnormal{DAF}}
%\newcommand{\ii}{\textnormal{i}}
\newcommand{\id}{\mathrm{d}\mkern1mu}
\newcommand{\ii}{\mathrm{i}\mkern1mu}
\newcommand{\nukg}{~\unit{\kilogram}\xspace}
\newcommand{\nustif}{~\unit{\newton\per\metre}\xspace}
\newcommand{\nudamp}{~\unit{\newton\second\per\metre}\xspace}
\newcommand{\nugpa}{~\si{\giga\pascal}\xspace}
\begin{document}
\onehalfspacing  % enable 1.5× spacing
\begin{frontmatter}
\title{A separation-of-variable dynamic stiffness matrix method for the free vibration of orthotropic rectangular thin plates}
\author[1]{Shiyi Mei}\ead{shiyi.mei1@monash.edu}
\author[1]{Colin Caprani\corref{cor1}}\ead{colin.caprani@monash.edu}
\author[2]{Daniel Cantero}\ead{daniel.cantero@ntnu.no}
\cortext[cor1]{Corresponding author}
\affiliation[1]{organization={Department of Civil Engineering, Monash University},city={Melbourne},state={Victoria},country={Australia}}
\affiliation[2]{organization={Department of Structural Engineering, Norwegian University of Science \& Technology NTNU},city={Trondheim}, country={Norway}}
\begin{abstract}
	A separation-of-variable dynamic stiffness matrix (SOV-DSM) method is proposed for the free vibration analysis of an orthotropic rectangular thin plate with general homogeneous boundary conditions.  
	Firstly, the exact SOV solution satisfying Rayleigh’s principle is extended to handle arbitrary boundary conditions and then applied to develop closed-form dynamic stiffness formulations.
	Then, an enhanced Wittrick–Williams (W-W) algorithm is used to solve the eigenvalue problem rather than solving the highly nonlinear eigenvalue equations.
	The $J_0$ count problem involved in applying the W-W algorithm is addressed by providing an explicit and closed-form expression for the $J_0$ term based on the characteristics of the SOV solution.
	Furthermore, a numerical technique is provided to calculate the mode shape of the plate with any arbitrary boundary conditions.
	The accuracy of the proposed method is validated through numerical experiments by comparison with other analytical solutions.
\end{abstract}
\end{frontmatter}
\linenumbers
\section{Introduction}
Rectangular plates play an important role in various engineering fields, including civil, mechanical, and aerospace engineering \citep{biancolini2005approximate}. 
The free vibration of plates has been a fundamental research problem for over two centuries. 
The earliest exact solutions for this problem are the Navier \citep{navier1823extrait} and Levy \citep{levy1899equilibre} solutions, which require at least one pair of opposite edges to be simply supported or guided.
To solve problems with other boundary conditions, approximate solutions such as the Rayleigh–Ritz method \cite{leissa1973free,lam1989vibration,zhao2025free,jain2024bending,bagheri2024analysis}, the Galerkin method \cite{laura1967study,krysl1995analysis,ma2024analysis,ma2024error,pereira2025modified} and finite element method (FEM) \cite{zhao2024fast,yang2022analysis,paccola2016continuous,carrera2024finite,yang2025semi} have been widely applied. 
For most approximation methods, beam functions \cite{ding1996natural,cheung1999free,cheung1999free2,zhou2002vibrations}, polynomials \cite{bhat1985natural,xie2025efficient,celep2025elastic,madinier2025spatial,sun2023spectral}, trigonometric functions \cite{filipich2000arbitrary,kumar2004buckling,yessenbayeva2019calculation,he2021study,su2022vibration}, and their combinations \cite{li2004vibration,cheung2000vibrations,zhou2001vibrations,monterrubio2015proof,onyeka2022application} are commonly used as the assumed approximate functions.
The accuracy of these solutions depends on how well the assumed approximate functions represent the displacement of the plate.

Besides the approximation methods, several analytical methods have been developed over past decades, including the Kantorovich-Krylov method \cite{kantorovich1958approximate,kerr1968extension,kerr1968application,rostami2016free,huang2015interlaminar,shufrin2016semi,singh2020three}, the symplectic method \cite{zhong1995new,xing2009new2,lim2009new,shi2024symplectic,xu2024hamiltonian,shi2025free}, the separation-of-variable (SOV) method \cite{xing2009new,xing2009new2,xing2018overall,xing2020extended,xing2020improved,xing2022review,wang2021closed,wang2021extended,wang2019highly,wang2020highly}, the dynamic stiffness matrix (DSM) method \cite{banerjee1997dynamic,papkov2019dynamic,liu2024dynamic,boscolo2011dynamic,fazzolari2013exact,ghorbel2015dynamic,williams1995wave,banerjee2015dynamic,liu2016free,liu2024wavenumber,zhou2024wavenumber,liu2015exact}, and series expansion-based methods \cite{timoshenko1940theory,gorman2005free,mochida2010transient,wang2016symplectic,hu2021new,wang2025new,khov2009accurate,li2009exact,grigorenko2003discrete,wang2016improved,leng2022new,li2009finite,zhong2013free,tian2015integral,zhang2017bending,chen2023analytical,li2024novel,zhu2024sound,shao2025analytic}.
The series expansion-based methods include the superposition method \cite{timoshenko1940theory,gorman2005free,mochida2010transient,wang2016symplectic,hu2021new,wang2025new}, Fourier series method \cite{khov2009accurate,li2009exact,grigorenko2003discrete,wang2016improved,leng2022new}, the finite integral transform method \cite{li2009finite,zhong2013free,tian2015integral,zhang2017bending,chen2023analytical}, and other series methods \cite{li2024novel,zhu2024sound,shao2025analytic}. 
These methods represent the plate displacement in terms of an infinite series and mostly are capable of handling any general boundary conditions. 
However, sufficient truncation of the series is required to ensure the accuracy and convergence of the results, and the eigenvalue equation is generally difficult to express explicitly.
Therefore, solving the corresponding eigenvalue problem can be computationally expensive.

Despite being a powerful method for the dynamic analysis of plate assemblies, the FEM requires a sufficient number of elements and is computationally expensive to accurately capture higher-order modes.
Thus, the DSM method was developed as an accurate and efficient analytical approach to alternatively solve complex plate structures \cite{boscolo2011dynamic,fazzolari2013exact}. 
The DSM can be considered as an analytical FEM since the mode functions of the plate are expressed by analytical solutions, where Levy-type solution \cite{ghorbel2015dynamic} or components of Fourier series \cite{williams1995wave,banerjee2015dynamic,liu2016free,liu2024wavenumber,zhou2024wavenumber} are applied.
To avoid solving the cumbersome transcendental frequency equation directly, the Wittrick-Williams (W-W) algorithm \cite{wittrick1971general} is applied to the eigenvalue problem.
The W-W algorithm determines the lower and upper bounds of natural frequencies to arbitrary precision rather than solving the frequency equation directly. 
Thus, the DSM has the potential to be effectively and systematically solved using the W-W algorithm.
However, a critical part in applying the W-W algorithm is to determine all natural frequencies of the fully clamped structure within the interested frequency range, \textit{a priori}.
Strategies such as using a sufficiently fine mesh or including a sufficient number of terms in series expansions \cite{banerjee2015dynamic} can ensure that all fully clamped frequencies are accounted for, thereby maintaining the accuracy of the algorithm. 
However, these approaches are computationally expensive and complex, posing a significant obstacle to the wider adoption and application of the DSM method based on the W-W algorithm \cite{han2018improved,liu2022closed}.
To resolve the fully clamped plate problem, Liu and Banerjee \cite{liu2015exact} suggested that the frequencies can be indirectly obtained from the simply supported plate problem, where the Navier solution serves as the analytical solution. 
This idea provides a significant enhancement to the W-W algorithm, increasing the efficiency of applying DSM methods \cite{liu2024wavenumber,zhou2024wavenumber}. 
However, most solutions applied in these DSM methods are not explicit and closed-form, but are expressed in an infinite series form, where a sufficient number of truncation terms might be required to ensure accuracy.

Inspired by the Navier and Levy solutions, Xing and Liu \cite{xing2009new} proposed the SOV method, which provides concise and explicit eigensolutions.
The mode shape function has a separable form, $\phi(x) \psi(y)$, requiring only one $\phi(x)$ and one $\psi(y)$ for each mode order, allowing each eigenvalue equation to be explicitly expressed.
However, this SOV method is not suitable to deal with plates with free boundary conditions.
Therefore, an extended SOV method \cite{xing2018overall, xing2020extended} based on the Rayleigh quotient was proposed to accommodate plates with all four classical boundary conditions, i.e., simply supported, clamped, guided, and free.
Based on the Rayleigh quotient model, alternative iterative and improved SOV methods have been subsequently proposed \cite{xing2020improved}.
Although SOV methods provide concise closed-form analytical solutions, they require solving a specific set of highly nonlinear eigenvalue equations for each type of boundary condition.
However, even when considering only the four classic homogeneous cases, it becomes evident that 55 different boundary condition combinations exist for a rectangular plate, making the process tedious.

In this study, a separation-of-variable dynamic stiffness matrix (SOV-DSM) method is proposed to avoid solving transcendental eigenvalue equations in SOV methods.
It also fills the gap caused by the limitation of closed-form dynamic stiffness formulations available for existing plate structures.
Firstly, the SOV method is further extended to analyze the vibrations of plates with elastically restrained edges based on Rayleigh’s principle. 
This extended SOV solution is then employed to construct dynamic stiffness matrices that accommodate general homogeneous boundary conditions.
By taking advantage of both the SOV and DSM methods, an enhanced W–W algorithm is developed to solve the eigenvalue problem without directly solving the eigenvalue equations.
This enhanced approach resolves the challenge of determining fully clamped frequencies, also called the $J_0$ count problem, a well-known limitation in the application of the W-W algorithm. 
Two types of explicit and closed-form expressions for the $J_0$ term, based on specific supports such as simply supported, guided, or their combinations, are provided to improve the computational efficiency of the eigenvalue problem.
In addition, a novel numerical technique is proposed to compute the mode shape coefficients.

The paper is structured as follows: \Cref{sec:SOV-DSM} presents the SOV-DSM formulations derived from the extended SOV method, which accommodates elastically restrained boundary conditions.
In \Cref{sec:WWA}, an enhanced W-W algorithm that resolves the $J_0$ count problem is provided to solve the dynamic stiffness matrices. 
The explicit and closed-form expression for the $J_0$ term is provided. 
In addition, a novel numerical technique is proposed to compute the mode shape of the plate with arbitrary boundary conditions.
\Cref{sec:NR} validates the proposed method through numerical experiments by comparison under several boundary conditions.
Finally, the conclusions of this paper are presented in \Cref{sec:coclusion}.

\FloatBarrier
\section{Separation-of-variable dynamic stiffness matrix formulation}\label{sec:SOV-DSM}
In this section, the SOV-DSM formulations for the free vibration analysis of orthotropic rectangular thin plates are developed.  
\Cref{sec:Mathematical model} presents the mathematical model of the extended separation-of-variable method.  
Then, the SOV-DSM formulations in the $x$- and $y$-directions will be proposed separately in \Cref{sec:DSMx} and \Cref{sec:DSMy}, according to the boundary conditions of the plate.

\subsection{Mathematical model for separation-of-variable method}\label{sec:Mathematical model}
\begin{figure}[!htbp]
\centering
\resizebox{0.8\textwidth}{!}
{
	\input{figs/general_plate_tikz.tex}
}
\caption{\small The orthotropic rectangular plate with all edges elastically restrained.} 
\label{fig:platemode}
\end{figure}
Consider a thin orthotropic rectangular plate of length $2a$ and width $2b$, with all four edges restrained by vertical translational springs $k^v$ and rotational springs $k^r$, as shown in \Cref{fig:platemode}. 
The coordinate origin is located at the center of the plate.

@@

Following the same steps as for the \(x\)-direction, the Rayleigh principle is expressed in the form:  

@@

@@

		\delta U &= \omega_y^2 \, \delta T_0,

@@

from which the governing differntial equation in the $y$-direction for $\psi$ is obatained as:

@@

	\alpha_2 &= \frac{1}{\chi} \sqrt{ \sqrt{ \left( \frac{D_{12}J_2}{D_{22}J_1} - 2\frac{D_{66}J_3}{D_{22}J_1} \right)^2 - \frac{D_{11}J_4}{D_{22}J_1} + \frac{a^4 D_{11}}{D_{22}} \Omega_y^4 } + \frac{D_{12}J_2}{D_{22}J_1} - 2 \frac{D_{66}J_3}{D_{22}J_1}}, \label{eq:betay1}\\
	\beta_2 &= \frac{1}{\chi} \sqrt{ \sqrt{ \left( \frac{D_{12}J_2}{D_{22}J_1} - 2 \frac{D_{66}J_3}{D_{22}J_1} \right)^2 - \frac{D_{11}J_4}{D_{22}J_1} + \frac{a^4 D_{11}}{D_{22}} \Omega_y^4 } - \frac{D_{12}J_2}{D_{22}J_1} + 2 \frac{D_{66}J_3}{D_{22}J_1}}.\label{eq:betay2}

@@

@@

\section*{Remarks}
\addcontentsline{toc}{subsection}{Remarks}
\begin{enumerate}[label=(\roman*)]
	\item For the SOV methods, six independent unknown eigenvalues ($\alpha_1$, $\beta_1$, $\Omega_x$, $\alpha_2$, $\beta_2$, $\Omega_y$) need to be determined for each mode shape $\phi(\xi)\psi(\eta)$.  
	To solve these six variables, four eigenvalue equations are provided in \Cref{eq:alphax1,eq:alphax2,eq:betay1,eq:betay2}, while the remaining two transcendental eigenvalue equations are obtained from the determinants of $\mathbf{R}_x$ in \Cref{eq:ABx1} for the $x$-direction and $\mathbf{R}_y$ in \Cref{eq:ABy1} for the $y$-direction, respectively.
	For classical boundary conditions such as simply supported (S), clamped (C), guided (G), and free (F) edges, these two transcendental equations have been simplified and summarized in previous works \citep{xing2020extended,xing2020improved,xing2022review}.  
	However, the process of selecting appropriate transcendental equations for the existing 55 different boundary-condition combinations remains tedious, and further simplification becomes challenging when non-classical boundary conditions are applied.
	By simply setting the translational springs ($k^v$) and rotational springs ($k^r$) along the $x$- or $y$-direction edges to either zero or infinity, the classical boundary conditions can be recovered.  
	In this case, the transcendental equations obtained from the determinants of $\mathbf{R}_x$ and $\mathbf{R}_y$ are identical to those summarized in the existing SOV methods.

	\item In the existing SOV methods, numerical techniques such as the Newton–Raphson
	method, homotopy method and optimization methods method have been applied to solve these six unknown variables.  
	Although all six eigenvalue equations can be solved simultaneously, the computation is very expensive and convergence issues may arise \citep{wang2021extended,wang2021closed}.  
	Therefore, an iterative procedure \cite{xing2018overall,wang2021extended,wang2021closed,wang2019highly,wang2020highly} is preferred, in which three eigenvalue equations corresponding to ($\alpha_1$, $\beta_1$, $\Omega_x$) and the remaining three corresponding to ($\alpha_2$, $\beta_2$, $\Omega_y$) are solved separately.
	In fact, by substituting \Cref{eq:alphax1,eq:alphax2} into the eigenvalue equation corresponding to $\mathbf{R}_x$ and \Cref{eq:betay1,eq:betay2} into the eigenvalue equation corresponding to $\mathbf{R}_y$, two highly nonlinear eigenvalue equations, involving only one variable $\Omega_x$ and $\Omega_y$, respectively, can be obtained. 
	However, these equations are difficult to solve using conventional numerical methods.
	To address this issue, the dynamic stiffness matrices $\mathbf{K}_x(\Omega_x)$ and $\mathbf{K}_y(\Omega_y)$ are developed, whose determinants are identical to the eigenvalue equations corresponding to $\mathbf{R}_x(\Omega_x)$ and $\mathbf{R}_y(\Omega_y)$, respectively.
	In this case, the dynamic stiffness matrices $\mathbf{K}_x(\Omega_x)$ and $\mathbf{K}_y(\Omega_y)$ can be solved efficiently using the W–W algorithm, which enhances computational robustness and ensures that no frequencies are missed.
\end{enumerate}
